\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-05 smurf_paper.tex
re-write of intro to section 3 done

%------------------------------------------------------------------------------
\section{The SMURF Algorithm}
\label{sec:algorithm}
%------------------------------------------------------------------------------

As described in the introduction, the approach taken by SMURF to
reducing \scuba\ data is to model (and remove) noise sources, and to
iterate this process along with estimates of the map. This sort of
algorithm has been used successfully in the past with instruments
including SCUBA \citep{1999ASPC..172..171J}, SHARC-2
\citep{kovacs2008}, and BOLOCAM \citep{aguirre2010}, and is capable of
recovering modest angular scales up to of order the footprint of the
array (depending, of course, on the noise in the data). The main
benefit of this approach over CMB-style maximum-likelihood methods
(which are generally superior at recovering structure on all scales
available in the data) is that the computing resources are
significantly smaller, both in terms of memory usage and execution
time. Given the massive volume of \scuba\ data produced in a single
observation (a rate of approximately 10 GB\,hr$^{-1}$ in each focal
plane), iterative techniques were the only viable option for providing
real-time feedback at the telescope, and for automatically reducing
near-science grade maps on a timescale comensurate with the
observation times using standard high-end desktop computers.  We do
note, however, that recent advances in direct inversion methods such
as SANEPIC \citep{patanchon2008} may in fact be tenable for \scuba\
data. A future goal of the software team is to investigate meshing
SMURF (perhaps as a cleaning step and initial condition) with a
SANEPIC-like algorithm.

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework with a simple (though extensive) interface for user
configurability. Though written in C, \smurf\ uses an object-oriented
style of handling the

As we have seen in Sections~\ref{sec:bolosignal}--\ref{sec:pca}

In practice, the model generally adopted for low-frequency noise
\scuba\ data is a single common-mode (average signal across all
working bolometers in a given focal plane), with most remaining noise
sources being removed by a high-pass filter. Many variations on this
were attempted throughout the commissioning phase, but as the PCA in
Secion~\ref{sec:pca} illustrates, less significant noise sources are
both transient and particularly difficult to model.

This section gives an overview of the SMURF algorithm, including data
cleaning and other pre-processing steps (Section~\ref{sec:dataprep}),
the iterative estimation of the map and noise model components
(Section~\ref{sec:components}), and concludes with a discussion of
convergence (Section~\ref{sec:converge}).

@@

@@

%-------------------------------------------------
\subsection{Model components}
\label{sec:components}
%-------------------------------------------------

@@

%-------------------------------------------------
\subsection{Data preparation}
\label{sec:dataprep}
%-------------------------------------------------

>>>>

%------------------------------------------------------------------------------
\section{Production of maps}
\label{sec:algorithm}
%------------------------------------------------------------------------------

As described in the introduction, the approach taken by SMURF to
reducing \scuba\ data is to model (and remove) predominantly
low-frequency noise sources that are correlated amongst detectors, and
to iterate this process along with estimates of the map. This sort of
algorithm has been used successfully in the past with instruments
including SCUBA \citep{1999ASPC..172..171J}, SHARC-2
\citep{kovacs2008}, and BOLOCAM \citep{aguirre2010}, and is capable of
recovering modest angular scales up to of order the footprint of the
array (depending, of course, on the noise in the data). The main
benefit of this approach over CMB-style maximum-likelihood methods
(which are generally superior at recovering structure on all scales
available in the data) is that the computing resources are
significantly smaller, both in terms of memory usage and execution
time. Given the massive volume of \scuba\ data (produced at a rate of
approximately 10 GB\,hr$^{-1}$ in each focal plane), iterative
techniques were the only viable option for providing real-time
feedback at the telescope, and for automatically reducing near-science
grade maps on a timescale comensurate with the observation times,
using single high-end desktop computers.  We do note, however, that
recent advances in direct inversion methods such as SANEPIC
\citep{patanchon2008} may in fact be tenable for \scuba\ data. A
future goal of the software team is to investigate meshing SMURF
(perhaps as a cleaning step and initial condition) with a SANEPIC-like
algorithm.

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework, and a configurable user interface. To achieve this goal,
while minimizing development time, it was decided to build SMURF as a
Starlink package \citep{2009ASPC..411..418J}, which provides access to
a large suite of libraries (including a commanding and messaging
interface, astrometric coordinate conversions and generation of
standard WCS information, file formats etc.). Furthermore, Starlink is
already used extensively at the Joint Astronomy Centre (host of the
JCMT) for many other systems, which helps with
interoperability. Though originally written primarily in FORTRAN, many
of the core Starlink libraries are now ported into native C, or else
have a C interface. It was therefore decided to develop SMURF in C
(rather than C++ which would have required adding further dependencies
to Starlink), although we have taken an object-oriented approach. For
example, all of the data for a given noise model is stored in a C
structure, and there is a standard interface for all functions that
handle models (member data and functions for the class,
respectively). In this way it is easy to extend SMURF with new
models. Parallelisation is incorporated in the most time-consuming
low-level routines using threads (e.g. performing Fast Fourier
Transforms, re-gridding the data), usually handling either independent
blocks of bolometers, or blocks of time, in each thread, depending on
the nature of the calculation. \textbf{... something about scalability
...}

@@

As we have seen in Sections~\ref{sec:bolosignal}--\ref{sec:pca},
\scuba\ data are dominated at low frequencies ($\lsim$2\,Hz) by highly
correlated signals. In addition to this $1/f$ noise, there is also
uncorrelated white noise, and finally residual correlated (but
non-astronomical) signals that appear at all frequencies with more
complicated relationships amongst groups of bolometers
(e.g. Fig.~\ref{fig:pca}). While the low-frequency noise can be
significantly reduced using simple common-mode removal (subtracting
the average signal from all bolometers at each time step), these
remaining correlated noise patterns are much more difficult to
model. Ultimately, a very simple approach has been taken to handling
these two contaminating sources, for which the main steps in a typical
reduction are shown in Fig.~\ref{fig:dimm}.

First, the raw data are passed through a pre-processing stage which
corrects some of the more significant glitches, applies flat-field
corrections etc. Next, the iterative process begins (dashed box). Most
of the low-frequency noise is removed using common-mode
subtraction. Next, the extinction correction is applied (it is done at
this stage rather than during pre-processing to avoid multiplying
large atmospheric signals by any small errors in the flatfield
correction). At this stage the data resemble the grey bolometer traces
in Fig.~\ref{fig:bolos_mix}, with some residual baseline drifts still
visible. These drifts are removed using a high-pass filter implemented
with Fast Fourier Transforms (FFTs). The residual signal is then
re-gridded to estimate the map. Finally, the map is used to estimate
and remove the astronomical signal contamination in the bolometer
data, leaving a relatively clean data set in which to measure the
white noise level of each bolometer. Since the common-mode and
filtering stages will have introduced ringing in the map in the
vicinity of bright sources, the entire process is iterated. Each
component in the dashed box of Fig.~\ref{fig:dimm} is re-calculated in
sequence. In this way, the second time the common-mode is calculated,
for example, most of the bright astronomical sources in the data have
already been removed in the previous iteration when the map was
estimated and subtracted, reducing the amount of ringing in the map.

In Sections~\ref{sec:dataprep} and \ref{sec:components} we describe
the data pre-processing steps and iterative algorithm in
detail. Section~\ref{sec:converge} explores convergence tests and
degeneracies between model components.

@@

%-------------------------------------------------
\subsection{Iterative model calculation}
\label{sec:components}
%-------------------------------------------------

@@

%-------------------------------------------------
\subsection{Data pre-processing}
\label{sec:dataprep}
%-------------------------------------------------
